---
title: "Basic Model formulation"
author: "Florian D. Schneider"
date: "Tuesday, November 13, 2014"
output:
  pdf_document:
    fig_height: 3.5
    fig_width: 4
  html_document:
    keep_md: yes
    self_contained: no
geometry: margin=1.4in
fontsize: 12pt
---
<!--

bibliography: cas02.bib
csl: amnat.csl
-->

This is supposed to become an interactive model representation of the pair-approximation model. 

The default form of the differential equation describing vegetation cover is 

$$ \frac{dV}{dt} = G(V) - C(V) $$

with growth function $G$ and a function $C$ which represents consumption by herbivores. Formally, intrinsic mortality could be included in both terms. In the formulations of @noy-meir75 and @vandekoppel97 it is assumed to be covered by the logistic growth function.

It might become important in the pair-approximation model, where intrinsic plant mortality is included as a proportional loss of mature plants.

```{r, echo=FALSE}
# Default plotting function

defplot <- function(...) plot(..., 
            type = "l", 
            bty = "l", cex = 0.7, las = 1, 
            family = "Times",
            xlim = c(0,1), ylim = c(0,0.2), 
            xaxs = "i", yaxs = "i", 
            xaxp = c(0,1,2), yaxp = c(0,0.2,2), 
            xlab = "vegetation cover", ) 
```


## logistic growth

$$ G(V) = rV(1-V/K)$$

```{r, echo=FALSE}
G <- function(P, r = 0.5, K = 0.9) r*P*(1-P/K)
rho <- seq(0,1,length = 100)

defplot(rho, G(rho), ylab = "growth")

```


## functional response


$$ C(V) = c_\mathrm{max} \frac{V}{V+T_h} $$

```{r, echo=FALSE}
C <- function(P, c_max = 0.1, T_h = 0.05) c_max*P/(P+T_h)
rho <- seq(0,1,length = 100)

defplot(rho, C(rho), ylab = "plant mortality")

```


```{r, echo=FALSE}

defplot(rho, G(rho), ylab = "growth")
lines(rho, C(rho, c_max = 0.1, T_h =0.3))
lines(rho, C(rho, c_max = 0.1))
```


